\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-22 chapters/bci.tex
Add decoders

Locked-in -\> communicate
Paraplegic -\> walk again
Imagine that
tetraplegic, locked-in syndrome
Regaining capabilities and improving quality of life
Start from wolpaw examples, for spelling
Cite State of the art examples (walk again, speech avatar Metzger, handwriting
willet, neuralink)

\todo{put focus on communication BCI}

@@

\fullpagefigct{figures/bci/bci_loop.pdf}{The \ac{bci}loop.
The user interacts with the \ac{bci} through a \ac{bci} paradigm, in this case involving
visual stimulation. Neural signals are recorded, and neurophysiological features
related to the paradigm are extracted. Using machine learning, a decision can
then be made based on these features, which can be presented back to the user.
In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{fig:bci-loop}

@@

The previous points to another component of a successful \ac{bci}: the specific
brain signals related to our paradigm need to be identified within the recorded
activity.
This is commonly referred to as \emph{feature extraction}
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own electrical signals.
On the other hand, the brain activity we are interested in exploiting often
originates from a specific functional system within the brain, and its activity
might not be easily discerned from all other neural and environmental activity,
akin to trying to listen to a single speaker in a noise environment where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out, but to obtain
a good measure reflecting only the activity that is relevant for a given \ac{bci}
problem.
This problem is called \emph{decoding} and is often solved through supervised
machine learning.
Section~\label{sec:bci-decoding} takes a closer look at some common and
state-of-the art techniques.
A machine learning algorithm can use some training data for which the performed
tasks are known, and learn to recognize these tasks in new data.
The training data can be either obtained from the \ac{bci} user themselves or from
other users.
In the first case, the user would perform a short calibration session before
using the \ac{bci}, where they are instructed to perform a known task.
This calibration session can be eliminated by training the decoder on
preexisting labeled data from other sessions and users, but this is harder due
to the variability between subjects and sessions.\todo{make shorter, move part
to section}

Finally, the loop can be closed by coupling this output can be coupled to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.\todo{put focus on communication BCI}
On the other hand, the actions performed by the \ac{bci} can themselves influence
the user's brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as an auditive prosthesis.
More indirectly modulation can be achieved by sensory stimulation.
This can involve specific, paradigm-related sensory input, like haptic feedback when a
movement \ac{bci} detects an intended action, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Due to neuroplasticity, the brain's ability to adapt, this can have a positive
impact on \ac{bci} performance and opens the pathway to rehabilitation applications.

@@

\Ac{eeg} is a de facto non-invasive electrophysiology standard, in clinical
neurology practice as well as in \ac{bci} research.
Developed in 1924, it is cheap, relatively practical and can offers a
temporal resolution.

outside skull, environmental/electrical noise, volume conduction activity from
everywhere in the brain.

cheap and practical, non-invasive, high-temporal resolution, but noise, suffers from volume
conduction hence low spatial resolution
what does ac{eeg} measure?

cost, ease of use

\section{\ac{bci} Paradigms}
\label{sec:bci-paradigms}
\fullpagefig{figures/bci/paradigms.eps}{caption}{fig:bci-paradigms}
\todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
paradigms}
passive:
also monitoring for diagnostic purposes
emotion recognition
neurofeedback

@@

somatosensory:

@@

visual paradigms:
SSVEP, cVEP, mVEP, VEP/P300/Oddball

\section{Decoders}
\label{sec:bci-decoders}
evaluated on
information transfer rate,
roc auc,
accuracy

Calibration, transfer learning, riemannian geometry, (multi)-linear techniques,
paper fabien lotte

active/reactive/passive + infographic

\section{A case study: the visual oddball ERP paradigm}
\subsection{stimulation}
\subsection{The event-related potential}
\todo{figure of P3 matrix}
P300 and other components
\todo{figure of ERP components}
\subsection{Decoding ERPs}
tLDA, beamforming
Linear decoding techniques

>>>>

\todo{
Locked-in -\> communicate
Paraplegic -\> walk again
tetraplegic, locked-in syndrome
Regaining capabilities and improving quality of life
Start from wolpaw examples, for spelling
Cite some catchy state of the art examples (walk again, speech avatar Metzger, handwriting
willet, neuralink)
}
\todo{put focus on communication BCI}

@@

\fullpagefigct{figures/bci/bci_loop.pdf}{The \ac{bci} loop.
The user interacts with the \ac{bci} through a \ac{bci} paradigm, in this case involving
visual stimulation. Neural signals are recorded, and neurophysiological features
related to the paradigm are extracted. Using machine learning, a decision can
then be made based on these features, which can be presented back to the user.
In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{fig:bci-loop}

@@

The previous points to another component of a successful \ac{bci}: the specific
brain signals related to our paradigm need to be identified within the recorded
activity.
\todo{preprocessing}
This is commonly referred to as \emph{feature extraction}
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own electrical signals.
On the other hand, the brain activity we are interested in exploiting often
originates from a specific functional system within the brain, and its activity
might not be easily discerned from all other neural and environmental activity,
akin to trying to listen to a single speaker in a noise environment where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out, but to obtain
a good measure reflecting only the activity that is relevant for a given \ac{bci}
problem.
This problem is called \emph{decoding} and is often solved through supervised
machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder.
Section~\label{sec:bci-decoding} takes a closer look at some common decoding
techniques.

Finally, the loop can be closed by coupling this to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.\todo{put focus on communication BCI}
On the other hand, the actions performed by the \ac{bci} can themselves influence
the user's brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as an auditive prosthesis.
More indirectly modulation can be achieved by sensory stimulation.
This can involve specific, paradigm-related sensory input, like haptic feedback when a
movement \ac{bci} detects an intended action, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Due to neuroplasticity, the brain's ability to adapt, this can have a positive
impact on \ac{bci} performance and opens the pathway to rehabilitation applications.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard, in clinical
neurology practice as well as in \ac{bci} research.
Developed in 1924, it is cheap, relatively practical and can offers great
temporal resolution up to thousands of Hz.
As it is easily applicable on the outside of the scalp, many practical
consumer grade \ac{eeg} headsets exist in the form (often wireless) electrode
caps or helmets.
Current consumer systems can feature dry electrodes, but clinical and
research systems often use electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

The major drawback of \ac{eeg} is its poor spatial resolution.
The electrodes are large and sit on the outside of the scalp, relatively far
away from the cortex.
Furthermore, the layers of skull and cerebrospinal fluid in between can
attenuate the signal and cause volume conduction, which results in electrodes
measuring a mixture of activity from sources elsewhere in the brain.
In addition to picking up brain activity other than that of the region of
interest, \ac{eeg} also records other artifacts, from the
environment, the power supply, nearby electrical equipment or the user's muscle
movements.
Combined, this means that \ac{eeg} is inherently noisy.

Despite its noisy nature \Ac{eeg} is the recording methodology of choice for
our \ac{bci} because of its wide acceptance by patients.
To counteract the noise present in our recording, we must on the one hand evoke
strong, informative brain signals using a suited \ac{bci} \emph{paradigm}, and on th
other hand pick a suited \emph{decoder} which can isolate the signal of
interest and filter out the noise.

\section{\ac{bci} Paradigms}
\label{sec:bci-paradigms}
\fullpagefig{figures/bci/paradigms.eps}{A \ac{bci} paradigm defines which brain
activity will be used for control. They can be categorized in \emph{active},
\emph{reactive} and \emph{passive} paradigms, depending on their reliance on
external stimulation and active participation required from the user.
Adapted from~\textcite{Muehl2014}.
}{fig:bci-paradigms}
\todo{check premissions}
\todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
paradigms}
passive:
also monitoring for diagnostic purposes
emotion recognition
neurofeedback
error potentials
cannot easily be used for fast communication
synchronous <-> asynchronous

\cite{Xu2021}

@@

somatosensory evoked potentials (cite jimmy)

@@

visual paradigms:
more performant since it allows us to use our visual system, the most adapted
sensory system in humans
SSVEP, cVEP, mVEP, VEP/P300/Oddball
high ssvep itr

hybrid

\section{Decoders}
\label{sec:bci-decoders}

After extracting features, machine learning algorithms can perform \ac{bci}
decoding.
This is usually based on
some training data for which the performed tasks are known, by recognize the
activity evoked by the task in recorded data.
The training data can be either obtained from the \ac{bci} user themselves or from
other users.
In the first case, the user would perform a short calibration session before
using the \ac{bci}, where they are instructed to perform a known task.
This calibration session can be eliminated by training the decoder on
preexisting labeled data from other sessions and users, but this is harder due
to the variability between subjects and sessions.\todo{make shorter, move part
to section}

\todo{regression/classification}
\todo{for communication, we want classification, focus on classification
further down the line}

If the goal is to eliminate the per-session calibration, one can also use a
decoder pre-trained on an existing dataset of the same task.
This is complicated by large variability in measured brain response between subjects
and even between different sessions within the same
subject~\cite{Guger2009,Saha2020}.
Therefore, pre-trained decoders must either be trained on a very large, diverse
sample of subjects, or some form of transfer learning must be applied.
Nevertheless, in practice pre-trained models or models using transfer learning often still
require some per session fine-tuning, which still necessitates some
calibration.
Instead, rather opt for keeping the calibration time as short as possible by using
an algorithm that can learn from very few training samples.
This can work well, but requires strong regularization~\cite{VanDenKerchove2022}.

\textcite{Lotte2018, Xu2021} present a relatively recent overview of state-of-the art
decoder algorithms.
Classic linear methods, such as \ac{csp} feature extraction for
motor imagery~\cite{Park2017} and \ac{cca} for \ac{ssvep}~\cite{Nakanishi2017},
and \ac{lda} for \ac{erp} classification~\cite{Sosulski2022} still perform
relatively well, given some regularizing constraints or extensions.
Multi-linear techniques exploiting the tensor structure of neural signals are
also promising.
Riemannian Geometry~\cite{Barachant2014} is a popular and robust new strategy,
Furthermore, they lend themselves well to application like adaptive
learning~\cite{Benaroch2021} or transfer learning~\cite{Zanini2017}
Riemannian Geometry classifiers are often considered the current
state-of-the-art.
Finally, deep learning\todo{Bhuvaneshwari2021} is also sometimes considered, albeit only when large
training datasets are available.
If decoders tailored to a specific user that keep the calibration session as
short as possible are of interest, not enough training data is available to
properly train a deep learning model.

Usually, a decoder makes a prediction for a given \emph{trial} while operating
the \ac{bci}.
A trial is the smallest unit on which a selection decision can be made, for
instance be one imagined movement, or one repetition of flashing all targets
in a visual \ac{erp} \ac{bci}.
Some valuable metrics to asses decoder performance in the classification case
are accuracy,
concept of trial
accuracy

single erp responses: roc auc, higher roc auc usually results in higher
accuracy

Finall, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can
be calculated as
\begin{equation}
	\text{ITR} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of
different options that can be selected per trial, $P$, the selection accuracy
of the decoder, and $Q$, the number of trials per second.
The parameters $N$, $P$ and $Q$ of this formula give us some insight in the
building blocks a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\todo{enumerate}
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a
broad range of informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for
decoding, and
\item increase $Q$ by selecting a paradigm and allows fast stimulation or
responses.
\end{enumerate*}

\section{A case study: the visual oddball ERP paradigm}
good choice because it scores high on the parameters above
\subsection{Stimulation}
\todo{figure of P3 matrix}
\todo{cite P300 literature, first example and current sota}
\subsection{The event-related potential}
P300 and other components
does not only rely on P300, but also heavily dependent on visual early
components, hence we'll refer to it under the more generall oddball ERP
paradigm
\todo{figure of ERP components}
\subsection{Preprocessing and feature extraction}
referencing, filtering, epoching, baselining
\subsection{Decoding ERPs}
tLDA, beamforming
Linear decoding techniques

### 2024-09-22 chapters/covert_align.tex
Add decoders

@@

@@

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention ERP (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
\ref{app:dataset} details the stimulation and recording procedure.
Each subject performed different VSA conditions (overt, covert and split),
illustrated in Figure~\ref{fig:interface}.
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The VSA conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

@@

\subsubsection{CVSA-ERP dataset}
EEG for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
(Compumedics Neuroscan, Australia) at 2048Hz and 62 Ag/AgCl active electrodes arranged in the
international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
Germany), with electrodes located at AFz and FCz as ground and reference respectively.
Using electrolyte gel, electrode impedances were brought below 5k$\Omega$.
Electrodes TP9 and TP10, used for off-line re-referencing, were directly
attached to the skin using stickers for better contact.
The power line frequency in Belgium is 50 Hz.
Participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
Canada) in non-fixation mode.

>>>>

\emph{This chapter has been published as part of~\textcite{VanDenKerchove2024}}

@@

\cite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert VSA compared to overt VSA.
This latency variability contributes to reduced covert VSA decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert VSA
decoding, but did not actually verify this hypothesis directly.
Additionally, \cite{hardiansyah2020single} developed a classifier for
covert VSA ERPs, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial ERP component latency
features on covert VSA inference yet did not attempt to correct the amplitude
features for these latencies.
\todo{Should we already mention Frenzel here?}
\cite{Frenzel2011} introduced a similar protocol to our split VSA
setting.
They showed that it is possible to perform split VSA and that, in this
case, visuospatial attention and gaze direction can separately be decoded using
classical ERP techniques.
To the best of our
knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
However, \cite{Frenzel2011} considered their interface only for the case
where the user actively intends to select both targets determined by the gaze
and the VSA.
In the split VSA setting considered in our work, we rather instruct the participant to ignore
the distractor and only attend the cued target, since we are interested in
decoding the visuospatial attention only.

@@

\subsubsection{CVSA-ERP dataset}

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention ERP (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
Each subject performed different VSA conditions (overt, covert and split),
illustrated in Figure~\ref{fig:interface}.
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The VSA conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

@@

EEG for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
(Compumedics Neuroscan, Australia) at 2048Hz and 62 Ag/AgCl active electrodes arranged in the
international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
Germany), with electrodes located at AFz and FCz as ground and reference respectively.
Using electrolyte gel, electrode impedances were brought below 5k$\Omega$.
Electrodes TP9 and TP10, used for off-line re-referencing, were directly
attached to the skin using stickers for better contact.
The power line frequency in Belgium is 50 Hz.
Participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
Canada) in non-fixation mode.
